Visualising metagenome differences

### 2023-11-21 analysis/metagenome.Rmd
dada

```{r}
colors <-c(microshades_palette("micro_blue", 4, lightest = FALSE), 
           microshades_palette("micro_purple", 4, lightest = FALSE),
           microshades_palette("micro_green", 4, lightest = FALSE),
           microshades_palette("micro_orange", 4, lightest = FALSE))
names(colors) <- c("Rhizobiales", "Clostridiales", "Desulfobacterales", "Chitinophagales", 
"Chromatiales", "Solirubrobacterales", "Xanthomonadales", "Acidimicrobiales", 
"Rhodobacterales", "Corallinales", "Ceramiales", "Alismatales", 
"Thiotrichales", "Flavobacteriales", "Remainder", "Unclassified")
```

```{r fig.height=6}
plot_orders_gtdb <- function(amph_pos_tax, group_remainder = TRUE) {
  temp <- amph_pos_tax |> 
    group_by(sample, Tissue, Sample_species,TotalMAGs) |> 
    mutate(Sample_species = trunc_str(Sample_species)) |> 
    mutate(Sample_species = paste0('*', Sample_species, '*')) |> 
    mutate(order = str_replace_all(order, 'o__','')) |> 
    count(order, .drop = FALSE) 
  if(group_remainder) {
    temp <- temp |> 
    mutate(counted_orders = case_when(n == 1 ~ 'Remainder',
                                      TRUE ~ order))
  } else {
      temp <- temp |> 
        mutate(counted_orders = order)
  }
  temp2 <- temp |> 
    select(-order) |> 
    group_by(sample, Tissue, Sample_species, n, counted_orders) |> 
    summarise(n = sum(n)) |> 
    filter(counted_orders != '' )
  # we need to count the unclassified MAGs that are not in the output
  group_presents <- temp2 |> group_by(sample) |> summarise(classifiedMAGs = sum(n))
  missings <- group_presents |> 
    left_join(temp |> 
                select(sample, TotalMAGs)) |> 
    unique() |> 
    mutate(unclassifiedMAGs = TotalMAGs - classifiedMAGs) |> 
    select(-c(classifiedMAGs,TotalMAGs)) |>
    rename(Unclassified = unclassifiedMAGs) |>
    pivot_longer(-c(sample, Tissue, Sample_species)) |> 
    rename(counted_orders=name, n=value)

  uniques <- c(setdiff(unique(temp2$counted_orders), c('Unclassified', 'Remainder')), 'Remainder', 'Unclassified')
  temp2 |> 
    mutate(counted_orders = factor(counted_orders, levels=uniques)) |> 
    mutate(sample = str_remove(sample, "[A-Z]+")) |> 
    mutate(sample = str_replace(sample, '31', '3')) |> 
    ggplot(aes(x = factor(sample, levels=rev(unique(sample))), y = n, fill=counted_orders)) +
    geom_col() + 
    theme_minimal() + 
    facet_wrap(~Tissue+Sample_species, scales = 'free_y', ncol=2) + 
    coord_flip() + 
    ylab('Number of contigs') + xlab('Replicate') +
    labs(fill = 'Order') +
    scale_y_continuous(breaks=pretty_breaks())  +
    # move legend position to bottom
    theme(legend.position = "bottom",
          strip.text.x = element_markdown()) +
    scale_fill_manual(values = colors)
}
plot_orders_gtdb(amph_pos_tax)
```

@@

    temp2 |>
      mutate(counted_orders = factor(counted_orders, levels = uniques)) |>
      mutate(sample = str_remove(sample, "[A-Z]+")) |>
      mutate(sample = str_replace(sample, '31', '3')) |>
      ggplot(aes(
        x = factor(sample, levels = rev(unique(sample))),
        y = n,
        fill = counted_orders
      )) +
      geom_col() +
      theme_minimal() +
      facet_wrap( ~ Sample_species + Tissue, ncol = 2, scales = 'free_y') +
      coord_flip() +
      ylab('Number of contigs') + xlab('Replicate') +
      labs(fill = 'Order') +
      scale_y_continuous(breaks = pretty_breaks())  +
      # move legend position to bottom
      theme(legend.position = "bottom",
            strip.text.x = element_markdown()) +
      scale_fill_manual(values = colors)
  }
plot_orders_cat(amph_pos_tax_cat)
```

@@

There we go :) Which MAGs are those?
```{r}
amph_pos_tax_cat |> filter(str_detect(Species, 'Acidihalobacter')) |> datatable()

```{r}
amph_pos_tax_cat |> filter(str_detect(Family, 'Ectothiorhodospiraceae'))|> datatable()
```
Are these in the GTDB classification?
```{r}
amph_pos_tax |> filter(str_detect(family, 'Ectothiorhodospiraceae'))|> datatable()
```
Nope!

@@

```{r}
count_diffs |> filter(`Posidonia australis` <100) |> datatable()
```

We hypothesise that the ACC precursor-existing genes in Amphibolis lead to different microbiomes. ACC is broken down by ACC deaminases, which are present in Amphibolis but not Posidonia. Let's see if we can find ACC deaminases in the metagenomes (*acdS*).

@@

```{r}
genessplit |> filter(gene == 'acdS') |> left_join(amph_pos_tax, by=c('sample', 'MAG'), multiple = 'any') |> 
  select(-c(Tissue.y, Sample_species.y)) |>
  rename(Tissue = Tissue.x, Sample_species = Sample_species.x) |> 
  plot_orders_gtdb(group_remainder = FALSE)
  #select(sample, MAG, domain:species)
```

>>>>

```{r}
colors <-c(microshades_palette("micro_blue", 4, lightest = FALSE), 
           microshades_palette("micro_purple", 4, lightest = FALSE),
           microshades_palette("micro_green", 4, lightest = FALSE),
           microshades_palette("micro_orange", 4, lightest = FALSE),
          microshades_palette("micro_brown", 4, lightest = FALSE),
           microshades_palette("micro_gray", 4, lightest = FALSE))
names(colors) <- c("Rhizobiales", "Clostridiales", "Desulfobacterales", 'Desulfarculales', "Chitinophagales", "SBR1031", "Chromatiales", "Solirubrobacterales", "Xanthomonadales", "Acidimicrobiales", "Arenicellales","Granulosicoccales",
"Rhodobacterales", "Corallinales", "Ceramiales", "Alismatales", 
"Thiotrichales", "Flavobacteriales", "UBA4575", "Remainder", "Unclassified")

```{r fig.height=6}
plot_orders_gtdb <- function(amph_pos_tax, group_remainder = TRUE, add_up=TRUE) {
  temp <- amph_pos_tax |> 
    group_by(sample, Tissue, Sample_species,TotalMAGs) |> 
    mutate(Sample_species = trunc_str(Sample_species)) |> 
    mutate(Sample_species = paste0('*', Sample_species, '*')) |> 
    mutate(order = str_replace_all(order, 'o__','')) |> 
    count(order, .drop = FALSE) 
  if(group_remainder) {
    temp <- temp |> 
    mutate(counted_orders = case_when(n == 1 ~ 'Remainder',
                                      TRUE ~ order))
  } else {
      temp <- temp |> 
        mutate(counted_orders = order)
  }
  temp2 <- temp |> 
    select(-order) |> 
    group_by(sample, Tissue, Sample_species, n, counted_orders) |> 
    summarise(n = sum(n)) |> 
    filter(counted_orders != '' )
  # we need to count the unclassified MAGs that are not in the output
  group_presents <- temp2 |> group_by(sample) |> summarise(classifiedMAGs = sum(n))
  missings <- group_presents |> 
    left_join(temp |> 
                select(sample, TotalMAGs)) |> 
    unique() |> 
    mutate(unclassifiedMAGs = TotalMAGs - classifiedMAGs) |> 
    select(-c(classifiedMAGs,TotalMAGs)) |>
    rename(Unclassified = unclassifiedMAGs) |>
    pivot_longer(-c(sample, Tissue, Sample_species)) |> 
    rename(counted_orders=name, n=value)

  if(add_up){
    # add hte missing numbers ot the main table
    temp2 <-rbind(temp2, missings)
  }
  temp2 <- temp2 |>  mutate(counted_orders = replace_na(counted_orders, 'Unclassified'))
  uniques <- c(setdiff(unique(temp2$counted_orders), c('Unclassified', 'Remainder')), 'Remainder', 'Unclassified')
  temp2 |> 
    mutate(counted_orders = factor(counted_orders, levels=uniques)) |> 
    mutate(sample = str_remove(sample, "[A-Z]+")) |> 
    mutate(sample = str_replace(sample, '31', '3')) |> 
    ggplot(aes(x = factor(sample, levels=rev(unique(sample))), y = n, fill=counted_orders)) +
    geom_col() + 
    theme_minimal() + 
    facet_wrap(~Tissue+Sample_species, scales = 'free_y', ncol=2) + 
    coord_flip() + 
    ylab('Number of MAGs') + xlab('Replicate') +
    labs(fill = 'Order') +
    scale_y_continuous(breaks=pretty_breaks())  +
    # move legend position to bottom
    theme(legend.position = "bottom",
          strip.text.x = element_markdown()) +
    scale_fill_manual(values = colors)
}
plot_orders_gtdb(amph_pos_tax)
```

@@

    temp2 |>
      mutate(counted_orders = factor(counted_orders, levels = uniques)) |>
      mutate(sample = str_remove(sample, "[A-Z]+")) |>
      mutate(sample = str_replace(sample, '31', '3')) |>
      ggplot(aes(
        x = factor(sample, levels = rev(unique(sample))),
        y = n,
        fill = counted_orders
      )) +
      geom_col() +
      theme_minimal() +
      facet_wrap( ~ Sample_species + Tissue, ncol = 2, scales = 'free_y') +
      coord_flip() +
      ylab('Number of MAGs') + xlab('Replicate') +
      labs(fill = 'Order') +
      scale_y_continuous(breaks = pretty_breaks())  +
      # move legend position to bottom
      theme(legend.position = "bottom",
            strip.text.x = element_markdown()) +
      scale_fill_manual(values = colors)
  }
plot_orders_cat(amph_pos_tax_cat)
```

@@

There we go :) Which MAGs are those?
```{r}
amph_pos_tax_cat |> filter(str_detect(Species, 'Acidihalobacter')) |> select(-c(lineage, `lineage scores`)) |> datatable()
```

```{r}
amph_pos_tax_cat |> filter(str_detect(Family, 'Ectothiorhodospiraceae'))|> select(-c(lineage, `lineage scores`)) |>  datatable()
```
Are these in the GTDB classification?
```{r}
amph_pos_tax |> filter(str_detect(family, 'Ectothiorhodospiraceae'))|> nrow()
```
Nope!

@@

```{r}
count_diffs |> filter(`Posidonia australis` <100) |> head() |> datatable()
```
Eeeh this is too unspecific. Let's look into specific differences.

## ACC deaminase
We hypothesise that the ACC precursor-existing genes in Amphibolis lead to different microbiomes. ACC is broken down by ACC deaminases, which are present in Amphibolis but not Posidonia. Let's see if we can find ACC deaminases in the metagenomes (*acdS*).

@@

TODO: The unknowns should have a color!!

```{r}
genessplit |> filter(gene == 'acdS') |> left_join(amph_pos_tax, by=c('sample', 'MAG')) |>
  select(-c(Tissue.y, Sample_species.y)) |>
  rename(Tissue = Tissue.x, Sample_species = Sample_species.x) |> 
  mutate(order = replace_na(order, 'Unclassified')) |>
  plot_orders_gtdb(group_remainder = FALSE, add_up = FALSE)
  #select(sample, MAG, domain:species)
```
